---
title: "Merge-Applications.R"
author: "Leah A. Wasser"
date: "April 1, 2016"
output: html_document
---

First, import the applications and clean out incompletes.

```{r}
# import data
# note these are the applications that are complete
apps <- read.csv("~/Documents/1_Summer-institute/inst-2016-apps-flags.csv", 
                 na.strings=c(-99, 0, "", "NA"),
                 stringsAsFactors = FALSE)


# remove some of the early columns that are not required
apps <- apps[,14:16]

# Then filter out applications according to status
# 1 = accepted, 2 - waiting list # 0 = not accepted
# each group will recieve a different email message.

```

## Accepted Applications

We accepted applications based upon their goals largely. If the goals fit with 
the content as we have defined it they were accepted in the first round. We 
also looked at their R and python skill level. However all applications submitted
atleast had the desired (basic) skill level. 

all applications were reviewed in google docs for ease of review. 
The scripts `mail-merge.R` and `app-review.Rmd` were used to sort all of the 
application information and produce clean looking pdf files. 

https://docs.google.com/spreadsheets/d/1rK6W3NxgN1PrMfjYAvl_I8PXo0tqeC4Wx5dLnu97vlU/edit

```{r filter-accepted, echo=FALSE}
# reader is nice as you don't have to worry about factors and it handles column
# classes better
library(readr)
library(plyr) # to join the df

# import data
apps.review <- read.csv("~/Documents/1_Summer-institute/application-review.csv",
                        stringsAsFactors = FALSE,
                        na.strings = c("NA"))

# join the review csv with the csv containing all applications
# note -- something odd is happening in the join where Sampath is coming in twice. probably
# a space but it doesn't matter given i am not accepting him anyway - incomplete app.
fin.apps.reviewed <- join(apps, apps.review, by = "lastName")

# now extract just the applications that were accepted to the institute
accepted.apps <- fin.apps.reviewed[(fin.apps.reviewed$Accept == 1),]

# remove NA values -- final accepted applicants
accepted.apps <- accepted.apps[!is.na(accepted.apps$Accept),]

# export final applicants
write_csv(accepted.apps, 
          "final-applicant-pool/accepted-all-apps.csv")
```

## Scholarships vs Non Scholarship

A few participants were awarded scholarships. We will send those participants
a slightly different acceptance email. Filter those out next

```{r scholarship}

# filter out those who requested scholarships
accepted.app.schol <- accepted.apps[accepted.apps$Scholarship==1,]

# write csv of those participants
write_csv(accepted.app.schol, 
          "final-applicant-pool/accepted-Schol-apps.csv")

# filter out those who requested scholarships
accepted.app.NOschol <- accepted.apps[accepted.apps$Scholarship==0,]

# write csv of those participants
write_csv(accepted.app.NOschol, 
          "final-applicant-pool/accepted-noSchol-apps.csv")

```


## Wait Listed  Applications

Some applications were wait listed. These individuals were often too advanced 
or topically a bit off what we plan to teach.

```{r not-accepted }

# now extract just the applications that were accepted to the institute
waitlist.accepted.apps <- fin.apps.reviewed[(fin.apps.reviewed$Accept == 2),]

# remove NA values -- final accepted applicants
waitlist.accepted.apps <- waitlist.accepted.apps[!is.na(waitlist.accepted.apps$Accept),]

# export final applicants
write_csv(waitlist.accepted.apps, 
          "final-applicant-pool/waitlist-apps.csv")


```

## Not Accepted This Year

Next, extract those who weren't accepted but had complete applications. We will
email them informing them they did not get in on the 4th.

```{r not-accepted }

# now extract just the applications that were accepted to the institute
not.accepted.apps <- fin.apps.reviewed[(fin.apps.reviewed$Accept == 0),]

# remove NA values -- final accepted applicants
not.accepted.apps <- not.accepted.apps[!is.na(not.accepted.apps$Accept),]

# export final applicants
write_csv(not.accepted.apps, 
          "final-applicant-pool/not-accepted-apps.csv")


```

