---
layout: post
title: "Visualize Precipitation & Stream Discharge Data in 
R to Better Understand - the 2013 Colorado Floods"
date: 2015-12-04
authors: [Leah A. Wasser, Mariela Perignon]
dateCreated: 2015-05-18
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
categories: [Coding and Informatics]
category: coding-and-informatics
tags: [R, time-series]
mainTag: GIS-Spatial-Data
scienceThemes: [phenology, disturbance]
description: "This lesson walks through the steps need to download and visualize
Precipitation Data and USGS Stream Discharge data in R 
to better understand the Drivers and Impacts of the 2013 Colorado floods."
image:
  feature: TeachingModules.jpg
  credit: A National Ecological Observatory Network (NEON) - Teaching Module
  creditlink: http://www.neoninc.org
permalink: /R/Boulder-Flood-Precipitation-Stream-Discharge-Data-R/
code1: Boulder-Flood-Data.R
comments: false
---

{% include _toc.html %}

Several factors contributed to extreme flooding that occured in Boulder, Colorado 
in 2013. In this lesson, we will explore two key variables including:

* precipitation (rainfall) data collected by ???NDCD?? 
* Stream discharge (or flow) that occured due to combination of the drought 
conditions and extreme rainfall.


<div id="objectives" markdown="1">

###Goals / Objectives
After completing this activity, you will:

* 

###Things You'll Need To Complete This Lesson
Please be sure you have the most current version of `R` and, preferably,
R studio to write your code.
####R Libraries to Install:

* **ggplot2:** `install.packages("ggplot2")`
* **lubridate:** `install.packages("lubridate")`

####Data to Download


<a href="#" class="btn btn-success"> 
#</a> 

The LiDAR and imagery data used to create the rasters in this dataset were 
collected over the <a href="http://www.neoninc.org/science-design/field-sites/harvard-forest" target="_blank" >Harvard</a> and 
<a href="http://www.neoninc.org/science-design/field-sites/san-joaquin-experimental-range" target="_blank" >San Joaquin</a> field sites 
and processed at <a href="http://www.neoninc.org" target="_blank" >NEON </a> 
headquarters. The entire dataset can be accessed by request from the 
<a href="http://www.neoninc.org/data-resources/get-data/airborne-data" target="_blank"> NEON 
website.</a>

####Recommended Pre-Lesson Reading


#### Lesson Series 

</div>

#R Libraries

We will be working with time-series data in this lesson so we will load the `lubridate`
library. We will use `ggplot2` to efficiently plot our data.

```{r load-libraries}

library(lubridate) #work with time series data
library(ggplot2)  #create efficient, professional plots
library(plotly) #create interactive plots

```


#Stream Discharge Data

Getting the data
http://waterdata.usgs.gov/nwis/


Map of all monitoring locations:
http://maps.waterdata.usgs.gov/mapper/index.html

Stream Guage Site
http://waterdata.usgs.gov/nwis/inventory?agency_code=USGS&site_no=06730200

Peak streamflow
http://nwis.waterdata.usgs.gov/usa/nwis/peak/?site_no=06730200

Auto retrievals
http://help.waterdata.usgs.gov/faq/automated-retrievals#RT

About Measurements:

http://water.usgs.gov/edu/measureflow.html

http://water.usgs.gov/edu/streamflow2.html

#note: it's unclear from this how the data were actually measursed given it seems
there are diferent sensors. Chuck bohall can probably help us some given he
worked at USGS.

##About The USGS Stream Monitoring Network

## Getting USGS Stream Gauge Data

Guage Data info:

USGS 06730200 BOULDER CREEK AT NORTH 75TH ST. NEAR BOULDER, CO

* We might include a map of where this guage is using leaflet.
* We can create the same map for the precip data location as well.

First - view the text file. Notice the first 25 lines are descriptive text rather
than data. Also notice that this file is separated by tabs rather than commas.

```{r import-discharge, echo=FALSE}
#SOURCE
#http://nwis.waterdata.usgs.gov/co/nwis/uv/?cb_00065=on&cb_00060=on&format=rdb&site_no=06730200&period=&begin_date=2013-01-01&end_date=2013-12-31
#import data
discharge <- read.table("precip-discharge/2013-discharge.txt",
                        sep="\t",
                        stringsAsFactors = FALSE,
                        skip=25,
                        header=TRUE)

#view first few lines
head(discharge)

#remove the first line from the data frame (which is a second list of headers)
boulderStrDis.2013 <- discharge[2:nrow(discharge),]

#view names
names(boulderStrDis.2013)

#rename headers so they are more meaningful
names(boulderStrDis.2013)[5] <- "disValue"

#view names
names(boulderStrDis.2013)
```

Let's have a look at the structure of our data. It appears as if the discharge 
value is a character class. This is likely because we had an additional row in our
data. Let's convert the discharge column to a numeric class. It appears as if 
`integer` will work given there are no decimal places in our data.

```{r adjust-data-structure }
#view structure of data
str(boulderStrDis.2013)

#convert column to integer
boulderStrDis.2013$disValue <- as.integer(boulderStrDis.2013$disValue)

class(boulderStrDis.2013$disValue)
str(boulderStrDis.2013)

```

#Converting Time Stamps
We have converted our discharge data to an integer format. However the time stamp
is still a character class. Let's convert it to a time class for efficient plotting,
next.

```{r convert-time }
#view class
class(boulderStrDis.2013$datetime)

#convert to date/time class - POSIX
boulderStrDis.2013$datetime <- as.POSIXct(boulderStrDis.2013$datetime)

#recheck data structure
str(boulderStrDis.2013)

#make sure there are no null values in our datetime field
sum(is.na(boulderStrDis.2013$datetime ))

```


Great - let's plot the data. We can use GGPLOT to plot.

```{r plot-flood-data }

ggplot(boulderStrDis.2013, aes(datetime, disValue)) +
  geom_point()

```


We can plot a subset of our data too. Let's specify the months directly around
the boulder flood - so August - October 2013.

```{r define-time-subset }

#just subset the plot to the month of Aug15-oct15
startTime <- as.POSIXct("2013-08-15 00:00:00")
endTime <- as.POSIXct("2013-10-15 00:00:00")

limits <- c(startTime,endTime)

```

#GGPLOT - 

One nice way to keep track of featured added to your plot in ggplot, is by first 
creating ... 

```{r plot-subset }
#plot the data - September-October
disPlot <- ggplot(data=boulderStrDis.2013,
       aes(datetime,disValue)) +
      geom_point() +
      scale_x_datetime(limits=limits)
      
#add title and labels
disPlot + theme(axis.title.x = element_blank()) +
          xlab("Time") + ylab("Stream Discharge CFS") +
          ggtitle("Stream Discharge - Station\n Boulder Creek 2013")


```

#Publish to Plot.ly

```{r plotly-discharge-data }
library(plotly)

#subset out some of the data - July-November
boulderStrDis.aug.oct2013 <- subset(boulderStrDis.2013, 
                        datetime >= as.POSIXct('2013-08-15 00:00',
                        tz = "America/New_York") & datetime <=
        as.POSIXct('2013-10-15 23:59', tz = "America/Denver"))

#plot the data - September-October
disPlot.plotly <- ggplot(data=boulderStrDis.aug.oct2013,
        aes(datetime,disValue)) +
        geom_point(size=3) 
      
#add title and labels
disPlot.plotly <- disPlot.plotly + theme(axis.title.x = element_blank()) +
          xlab("Time") + ylab("Stream Discharge CFS") +
          ggtitle("Stream Discharge - Boulder Creek 2013")

#view plotly plot in R
ggplotly()

#publish plotly plot to your plot.ly online account when you are happy with it
plotly_POST(disPlot.plotly)

```

#Explore Precipitation During the Flood

```{r import-precip }

precip.boulder <- read.csv("precip-discharge/2013-precip-data-640457.csv",
                           stringsAsFactors = FALSE)


#convert date time
precip.boulder$DATE <- as.POSIXct(precip.boulder$DATE,
                                  format="%Y%m%d %H:%M")

#clean up no data values
precip.boulder$HPCP[precip.boulder$HPCP==99999] <- NA

#plot the data - September-October
precPlot <- ggplot(data=precip.boulder,
      aes(DATE,HPCP)) +
      geom_bar(stat="identity") +
      scale_x_datetime(limits=limits) +
      scale_y_continuous(limits = c(0, 300))
      
#add title and labels
precPlot + theme(axis.title.x = element_blank()) +
          xlab("Time") + ylab("Precipitation (100th of an inch)") +
          ggtitle("Hourly Precipitation - Boulder - Station\n Boulder Creek 2013")

```


```{r daily-summaries }

library(plyr)
#add a day column in the data frame
precip.boulder$daily <- floor_date(precip.boulder$DATE, "day")

#summarize by day
dailyPrecip.boulder <- ddply(precip.boulder, "daily",summarise, x=sum(HPCP))

names(dailyPrecip.boulder)

#let's create a more meaningful column header for the precip value
names(dailyPrecip.boulder) <- c("day","prec_100In")

# view names
names(dailyPrecip.boulder)

#are there no data values?
sum(is.na(dailyPrecip.boulder))

#remove rows with no data
dailyPrecip.boulder.cln <- na.omit(dailyPrecip.boulder)

#plot the data - September-October
precPlot <- ggplot(data=dailyPrecip.boulder.cln,
      aes(day,prec_100In)) +
       geom_bar(stat="identity") +
      scale_x_datetime(limits=limits) +
      scale_y_continuous(limits = c(0, 800))
      
#add title and labels
precPlot + theme(axis.title.x = element_blank()) +
          xlab("Time") + ylab("Precipitation (100th of an inch)") +
          ggtitle("DAILY Precipitation - Boulder- Station\n Boulder Creek 2013")


```

## Units

It looks like our precipitation data are in 100th of an inch. Let's convert the 
data to inches - units that we might be a bit more accustomed to. And then let's
plot the data as a bar plot rather than a scatter plot.

```{r convert-units }


#convert to inches
dailyPrecip.boulder.cln$prec_in <- dailyPrecip.boulder.cln$prec_100In / 100

head(dailyPrecip.boulder.cln)

#plot the data - September-October
precPlot_2013 <- ggplot(data=dailyPrecip.boulder.cln,
      aes(day,prec_in)) +
      geom_bar(stat="identity") +
      scale_x_datetime() +
      scale_y_continuous(limits = c(0, 8))
      
#add title and labels
precPlot_2013 <- precPlot_2013 + theme(axis.title.x = element_blank()) +
          xlab("2013") + ylab("Precipitation (Inches)") +
          ggtitle("Daily Total Precipitation (Inches) - Boulder- Station\n Boulder Creek 2013")


precPlot_2013

#plot the data - September-October
precPlot <- ggplot(data=dailyPrecip.boulder.cln,
      aes(day,prec_in)) +
      geom_bar(stat="identity") +
      scale_x_datetime(limits=limits) +
      scale_y_continuous(limits = c(0, 8))
      
#add title and labels
precPlot <- precPlot + theme(axis.title.x = element_blank()) +
          xlab("Time") + ylab("Precipitation (inches)") +
          ggtitle("Daily Total Precipitation (Inches) - Boulder- Station\n Boulder Creek 2013")


precPlot
```

#Interactive Plots - Plot.ly

Let's turn our plot into an interactive plot.ly plot



```{r plotly-precip-data }

library(plotly)
#setup your plot.ly credentials
Sys.setenv("plotly_username"="your.user.name.here")
Sys.setenv("plotly_api_key"="your.key.here")

#subset out some of the data - July-November
dailyPrec.sub <- subset(dailyPrecip.boulder.cln, 
                        day >= as.POSIXct('2013-08-15 00:00',
                        tz = "America/New_York") & day <=
        as.POSIXct('2013-10-15 23:59', tz = "America/Denver"))


#create new plot
new <- ggplot(data=dailyPrec.sub, aes(day,prec_in)) +
  geom_bar(stat="identity") +
  xlab("Time") + ylab("Precipitation (inches)") +
  ggtitle("Daily Total Precipitation (Inches) - Boulder Creek 2013") 

#view plotly plot in R
ggplotly()

#publish plotly plot to your plot.ly online account when you are happy with it
plotly_POST(new)


```

