---
title: "Institute-Applications"
output:
  pdf_document:
    latex_engine: xelatex
---


```{r setup-functions, echo=FALSE}

library(readr)
library(stringr)
# build list function

fun_bldList<- function(a.string, cols) {
  # a.string <- "Q33_"
  # cols=1
  # build list of all options for question 19
  q.names <- grep(a.string, names(apps))
  # create list of responses for each language option
  element.list <- as.data.frame(t(apps[i, q.names]), stringsAsFactors = FALSE)
  # create a list of response questions for optional use
  # pull the first row which contains the question
  titles <- as.character(apps[1, q.names])
  
  # split titles at dash
  # this is causing issues now as there's no consistent way to break this up.
  #df.titles <- as.data.frame(strsplit(titles, "\\.\\.-"), stringsAsFactors = FALSE)
  df.titles <- str_split(titles, "-", n=2)
  df.titles <- do.call(rbind.data.frame, df.titles)
  #df.titles <- as.data.frame(str_split(titles, "-", n=2), stringsAsFactors = FALSE)

  # just pull the question title 
  # df.titles <- df.titles[2,]
  # names(df.titles) <- df.titles[1,]
  df.titles[,3] <- element.list[,1]
  
  if (cols==1){
    # remove NA values
    element.list <- element.list[!is.na(element.list)]
    return(element.list)
} else {
  # remove rows in the DF with NA
  #df.titles <- df.titles[!is.na(df.titles)]
  #df.titles[2,] <- names(df.titles)
  return(df.titles)
  }
}

```

```{r import-data, echo=FALSE}

# import data
apps <- read.csv("~/Documents/1_Summer-institute/inst-2016-apps-flags.csv", 
                 na.strings=c(-99, 0, "", "NA"),
                 stringsAsFactors = FALSE)

# filter out Q17 (goals) == NA
inCompleteApps <- apps[is.na(apps$Q17),]
apps <- apps[!is.na(apps$Q17),]

```

```{r populate-variables, echo=FALSE}
# implement data cleaning -- replace dashes with slash
# gsub("-", "/", apps)

# set all values of -99 to NA
apps[apps==-99] <- NA

# get contact info and background
firstName <- apps$Q4[i]
lastName <- apps$Q5[i]

email <- apps$Q6[i]
phone <- apps$Q7[i]
state <- apps$Q10[i]
city <- apps$Q8[i]
country <- apps$Q9_TEXT[i]

institution <- apps$Q12[i]
country <- apps$Q9_TEXT[i]
domain <- apps$Q14[i]
position <- apps$Q13[i]
position.text <- apps$Q13_TEXT[i]
position.other <- apps$Q15[i]

# goals
goals <- apps$Q17[i]


# create list of languages - question 19
language.list <- fun_bldList("Q19_", 1)
ext.data.list <- fun_bldList("Q21_", 1)
ext.RS.list <- fun_bldList("Q22_", 1)
rs.list.desc <- apps$Q23[i]

ext.meta.list <- fun_bldList("Q24_", 1)
file.formats.list <- fun_bldList("Q26_", 1)
document.list <- fun_bldList("Q25_", 1)

# this is throwing an error?

cost.ack <- apps$Q33[i]
funding.source <- apps$Q33[i]

funding.explanation <- apps$Q34[i]



# return a data frame so i can print the question text
data.freq.list <- fun_bldList("Q20_", 2)
spatial.tools.list <- fun_bldList("Q27_", 2)
r.comfort <- fun_bldList("Q28_", 2)
py.comfort <- fun_bldList("Q29_", 2)


```
 
## `r firstName` `r lastName`

**email:**  `r email`, **ph:**  `r phone`,

**Location:** `r city`, `r state`
**Country:** `r country`
 
**Affiliation:** `r institution`

**Position:** `r position`, `r if (!is.null(position.text)) { position.text }`
`r if (!is.na(position.other)) { position.other }`

**Domain:** `r domain`
 
****
## Goals

`r goals`
 
## Scholarship Request

`r cost.ack`

`r funding.source`

**Request For Scholarship:** `r funding.explanation`

**** 

## What `r firstName` Already Knows 

**Languages:** `r language.list` 

**Data Used:** `r ext.data.list` 

***

**Remote Sensing Used:** `r ext.RS.list`

**RS notes:** `r rs.list.desc`

***

**Metadata Used:** `r ext.meta.list`

**Familiar File Formats:** `r file.formats.list`

**Code Documentation:** `r document.list`



```{r print-knowledge, results='asis', echo=FALSE}
for (someData in (1:nrow(data.freq.list))){ 
  cat(paste0("* **", data.freq.list[someData,2], ":** ", data.freq.list[someData,3], "\n"))
  }


```

### Spatial Tools List:


```{r spatial-tools-knowledge, results='asis', echo=FALSE}
for (someData in (1:nrow(spatial.tools.list))){ 
  cat(paste0("* **", spatial.tools.list[someData,2], ":** ", spatial.tools.list[someData,3], "\n"))
  }
```

### Comfort in R:


```{r rr-knowledge, results='asis', echo=FALSE}
for (someData in (1:nrow(r.comfort))){ 
  cat(paste0("* **", r.comfort[someData,2], ":** ", r.comfort[someData,3], "\n"))
  }
```


### Comfort in Python:


```{r py-knowledge, results='asis', echo=FALSE}
for (someData in (1:nrow(py.comfort))){ 
  cat(paste0("* **", py.comfort[someData,2], ":** ", py.comfort[someData,3], "\n"))
  }
```



